Cheminformatics: The Computer Science of Chemical Discovery

### 2011-11-08 paper.tex
put in submitted version of the paper from Aaron

@@

\maketitle
\begin{abstract}
One of the most prominent success stories in all the sciences over the last decade has
  been the advance of bioinformatics: The interdisciplinary collaboration between
 computer scientists and molecular biologists led as one its major accomplishments to the sequencing of the human genome, where researchers connected biological concepts to the theory of string algorithms. However, despite this great
  success, few computer scientists are familiar with a related (and older!)
  discipline: cheminformatics, the use of computers to represent
 the structures of small molecules and analyze their properties.
  Cheminformatics has wide applicability, from the discovery of novel drug candidates to the optimization of
  certain physicochemical properties of small molecules. Until
 recently, the data and techniques employed in the cheminformatics field have been closely guarded
  secrets of companies whose financial success depended on being the first to produce
  the new ``miracle molecule.'' Only within the last decade---and as an effect of a change in mindsets, government-mandated policy changes and last but not least because of  chemists volunteering their time for an Open Science ``movement''---have researchers gained access to freely available software packages and databases of tens of
  millions of chemicals. Academic chemists now confront a variety of unsolved
  algorithmic problems that could not have been tackled a decade ago, but whose
  solutions are critical to research ranging from determining the behavior of small
  molecules in biological pathways, to finding a cure for malaria as well as many other neglected diseases.
\end{abstract}

\category{H.4}{Information Systems Applications}{Miscellaneous}
\terms{
\begin{itemize}
\item \textit{Bioassay} - A system measuring the biological activity of a chemical compound on a particular biological assay (with all its parameters).
\item \textit{Biodegradation} - is the chemical dissolution of materials by bacteria or other biological means.
\item \textit{Conformation and conformational isomerism} - In chemistry, conformational isomerism is a form of stereoisomerism in which the isomers can be interconverted exclusively by rotations about formally single bonds. Single possible states of compounds are called conformers.
\item \textit{Ligand (biochemistry)} - a substance that binds to a protein.
\item \textit{Ligand (coordination chemistry)} - an atom, ion, or functional group that donates one or more of its electrons through a coordinate covalent bond to one or more central atoms or ions.
\item \textit{Mass spectrometry} - Mass spectrometry (MS) is an analytical technique that measures the mass-to-charge ratio of charged particles. It is used for determining masses of particles, for determining the elemental composition of a sample or molecule, and for elucidating the chemical structures of molecules, such as peptides and other chemical compounds.
\item \textit{Metabolites} - Metabolites are the intermediates and products of metabolism. The term metabolite is usually restricted to small molecules (chemical compounds).
\item \textit{Proteochemometric modeling} - combining data mining for chemical compounds (ligands) and proteins within the same machine learning framework.
\item \textit{Protonation and protonation state} - In chemistry, protonation is the addition of a proton (H$^+$) to an atom, molecule, or ion. The different states (different number of H$^+$ or on different atoms) are called protonation states for a molecular compound (e.g. de-protonated, mono-protonated, di-protonated, or any variations of those states).
\item \textit{QSAR} - Quantitative Structure-Activity Relationships; methods to relate the structure and the biological activity of a chemical structure in a quantitative manner.
\item \textit{Reaction catalysis} - Catalysis is the change in rate of a chemical reaction due to the participation of a substance called a catalyst.
\item \textit{Receptor (biochemistry)} - in biochemistry, a protein molecule that receives and responds to a neurotransmitter, or other substance.
\item \textit{Regulation authorities} - In the US the Food and Drug Administration (FDA) is responsible for protecting and promoting public health through the regulation and supervision of e.g. dietary supplements, prescription and over-the-counter pharmaceutical drugs (medications), vaccines, biopharmaceuticals, medical devices, and many other health related items. The European Medicines Agency (EMA) is a European agency for the evaluation of medicinal products.
\item \textit{SMARTS} - SMiles ARbitrary Target Specification \\(SMARTS) is a language for specifying substructure patterns in molecules
\item \textit{Stereoisomer/stereoisomerism} - Stereoisomers are isomeric molecules that have the same molecular formula and sequence of bonded atoms (constitution), but that differ only in the three-dimensional orientations of their atoms in space.
\item \textit{Tautomerism} - Tautomers are isomers (same molecular formula but different structural formula) of organic compounds that readily interconvert by a chemical reaction called tautomerization.
\end{itemize}
}
\keywords{cheminformatics, chemoinformatics, graphs, chemistry, databases, knowledge management, open source, open data}

@@

Novel technologies in the life sciences churn out information at an ever
increasing rate---public data stores such as the one at the European Bioinformatics
Institute (EBI) contain on the order of 10 petabytes of biological information.
However, while biological information has been publicly available for decades,
the same has not been true for chemical information for very long. It was
not until 2004 that a large public small molecule structure repository (PubChem)
became freely available. This was followed by other databases, as discussed later
in this article. In the same vein, open source software and the publication of
algorithms have only recently become a main focus of the field
\cite{faulon2010}.

@@

\section{Bridging Cheminformatics \& Computer Science}

@@

While cheminformatics has a rather broad applicability, ranging from
agrochemical research and drug discovery to the design of novel
materials (since everything around us consists of 'molecules'), we have chosen to use drug discovery as the context of this
article due to its central relevance for human wellbeing. Given that the safety of drugs on the market is a key concern both for the general audience as well as researchers in pharmaceutical companies, the remainder of this article is structured around the
concept of ``risk minimization'' in drug discovery---the question of how to minimize the chances of a small molecule failing in the various research and development stages as a drug candidate, and how to successfully employed cheminformatics methods in these efforts.

\subsection{Databases \& Ontologies}
\label{sec:databases}

Most cheminformatics applications rely on large databases of chemical structures, 
their properties, and relationships to e.g. biological targets.
Organizing and maintaining such databases, as well as searching and clustering similar structures together,
are essential  to enable many scientific applications. Yet, each one of
those areas poses its unique computer science challenges.

There is considerable progress in open chemical and bioactivity databases like
ChEMBL and NCBI-PubChem, which are both freely available and contain in the order of millions (ChEMBL) and tens of millions of data points each (PubChem). Still, the integration
of chemical databases with each other remains a challenge, not only due to the
normalization problems in the chemical and bioactivity landscape, but also due
to sheer data volume. One problem is the annotation with the 'protein targets' a molecule binds to - here dozens of protein and gene identifiers exist, generally without the availability of a one-to-one mapping from each ontology, which makes merging databases often a very cumbersome exercise.

Given the requirement to 'make sense' of as much chemical data at hand as possible, we have to be able to search in complex data, for example multi-labelled molecular graphs, where graph labels can be different. Hence, we need to be able to capture
molecular graphs in a risk-bounded machine encoded way.

One main challenge in the cheminformatics field is that not one unique way exists to represent every molecule in a machine-readable form. This results in multiple encodings of
``the same'' molecule and its different tautomeric or protonation states (see glossary for an explanation of the technical terms). Checking whether two graphs represent the same chemical structure is
an expensive operation (the next section will discuss this in-depth);
thus, a representation of the structure that is unique and invariant
with respect to a particular graphical display of that chemical is needed for
identification. An early effort in this direction was the Simplified
Molecular Input Line Specification (SMILES), and specifically the
generation of a single, 'canonical' SMILES for a particular structure. The idea was to devise rules for a unique ordering of the atoms in a molecular structure, hence resulting in the same representation of the same molecule. However,
over the course of time, several
different SMILES implementations became available, with no universal
standard for generating a canonical SMILES. Nonetheless, the SMILES format remains one of the major representations of molecules used in practice today.

With each database using their own algorithm for molecular encoding,
automated exchange of data between different databases was hindered.
As more and more data became available online, the use of unique
structure-based identifiers became a pressing need, resulting in the IUPAC
International Chemical Identifier (InChI) to be developed. The
InChI is a non-proprietary, structured textual identifier for chemical
entities \cite{inchi}. InChI identifiers are not intended to be read and
understood by humans, but are useful for computational matching of
chemical entities. For rapid database lookups, the InChIKey is a
hashed key for the InChI, which has an invariant length of 14
characters. (Unfortunately, after implementation and publication of the hashing scheme it became apparent that in particular in case of stereochemically related molecules clashes of the InChI key take place in some, albeit rare cases.) Figure \ref{smiles} illustrates the SMILES, InChI and InChIKey for
lipoic acid.

%Figure {smiles}: The standard chemical graph illustration together with the SMILES,
%InChI and InChIKey codes for lipoic acid. Note that the SMILES is much more
%human-readable than the others.

The InChI is now widely used for matching identical chemical structures, but it
is still limited. For example, it cannot differentiate between
certain types of stereoisomers: Figure \ref{cistrans} illustrates two stereoisomers for which
the generated InChI is the same.

%Figure {cistrans}: The generated InChI is the same for the stereoisomers cisplatin and
%transplatin.

@@

Graph substructure matching is a variant of \emph{graph isomorphism detection},which is
widely believed to be computationally intractable [3]. To execute a graph isomorphism
search across a full chemical database of thousands or even millions of
structures is infeasible. For that reason, complex heuristics must be
used to drastically filter the candidate search space. Structure-based
\emph{fingerprints} are commonly used for this purpose. Fingerprints encode
characteristic features of a given chemical structure in a Boolean array, or
bitmap, and they can be implemented in a variety of ways. Common fingerprints are created by an algorithm which generates bitmap patterns for each atom in the structure; then each atom and its nearest
neighbors, including the bonds between them; then each group of atoms connected
by paths of increasing lengths up to the maximum implemented for the algorithm
(commonly around eight). In this way, characteristic parts of molecule are captured and the bitmap patterns are added together into the final fingerprint. If all bits in a query fingerprint are also present in the target
fingerprint of a stored database structure, only then is the structure subjected
to the computationally expensive subgraph matching algorithm. Bit operations are
very fast and independent of the number of atoms in a structure due to the fixed
length of the fingerprint.

Fingerprints also facilitate the rapid calculation of quantitative
structure-based similarity measures. An example measure of similarity is the
\emph{Tanimoto coefficient}, calculated as the ratio $T(a,b) = \frac{c}{(a+b-c)}$,
where $c$ is the count of bits on in the same position in both the two
fingerprints, $a$ is the count of bits on in the first structure, and $b$ is the
count of bits on in the second structure. Historically, this coefficient originated from the comparison of ecosystems around lakes, where a measure was needed to describe how similar those ecosystems are with respect to the animals, plants etc. they contain. However, it can be applied to every other system as well where features can be described in a binary manner. The Tanimoto coefficient varies in the
range 0.0 -- 1.0, with a score of 1.0 indicating that the two structures are either very
similar or even identical (i.e. their fingerprints are the same).

Structure-based similarity is however of
limited use in grouping of chemicals based on non-structural features, such as
shared bioactivity profiles. For this purpose, a large-scale and flexible
classification structure is needed for chemicals and their linked biological
information. One such answer to this need is provided in the form of \emph{ontologies}.
An ontology is a formal specification of entities and their relationships in a
domain of interest. It is structured around an underlying logical representation
such as Description Logics~\cite{baaderdl2007}. The logic-based representation allows
automated reasoning to derive new knowledge (inferences) from the knowledge
encoded, making the representation both compact and powerful.
Ontologies are already in widespread use across the biomedical domain, and in the cheminformatics field ontologies relating to the chemical structures themselves and the protein targets the molecules are active against are probably most common. A
similarity measure based on the information encoded in ontologies is termed
\emph{semantic similarity}, and such measures have already been applied to enhance
chemical classification in several problem areas \cite{couto2010}. Many open
questions still remain about the best way to combine the information encoded in
chemical structures with chemical ontologies \cite{hastingsowled2010}.

@@

Enumerating molecules is a combinatorial problem that has fascinated
chemists, computer scientists and mathematicians alike, for more than a
century. It is one important risk-reduction element in still being
able to encode closely-related molecules and for keeping the space
complexity of databases under (algorithmic) control. Indeed, while
trying to solve the problem of counting the isomers of paraffin
structures or counting substituted aromatic compounds, fundamental
principles in graph theory and combinatorics were developed by Cayley,
Polya and others. Even terms that are widely-used nowadays, like 'graph'
and 'tree', were originally coined in a chemistry context (Chapters 1 and
8 in \cite{faulon2010}). More than four decades ago, Joshua Lederberg,
Carl Djerassi, Edward Feigenbaum and others from Stanford University
started developing codes to explicitly enumerate molecules. Their
project, named DENDRAL, was funded by NASA with the goal of developing
an automated system to elucidate compounds from spectral data. While
studying this challenging problem, important concepts in computer
science were developed. DENDRAL is generally quoted in artificial
intelligence textbooks as a pioneer project that produced the first
expert system.

Enumerating molecules is not only an interesting academic exercise but has
practical utility as well. The foremost application of enumeration is \emph{structure
elucidation} of natural compounds such as metabolites. Metabolites are small molecules that can be found an organism (such as a human) and which in many cases indicate the state the organism is in - such as indicating a disease, or a fasted state, etc., and with a sufficiently large database the idea is to be able to detect the chemical composition of body fluids and to arrive at conclusions regarding the particular state of the organism at this point in time.
Ideally, the wishful bench chemist collects experimental data
from an unknown compound, the data is fed into a code, and the resulting
structure is returned. Although that streamlined process may not yield a unique
solution, there are commercial software products such as MOLGEN
(http://molgen.de) that can, for instance, list all structures matching a given
molecular formula, or mass spectrometry (MS) data. Another important application
of enumeration is \emph{molecular design}. Here the problem is to design compounds
(drugs, for instance) that optimize some physical, chemical, or biological
property or activity. Although less prolific and more recent than structure
elucidation, molecular design has introduced some novel solutions to molecular
enumeration. Finally, with the advent of combinatorial chemistry, molecular
enumeration has taken a central role as it allows computational chemists to
construct virtual libraries and test hypotheses, and it provides guidance in the
design of optimal combinatorial experiments.

The major difficulty with enumeration (apart from the size of the search space, which is thought to be in the order of 10^60 molecules for 30 heavy atoms) is that the \emph{in silico}
representation of molecules, the so-called molecular graphs, are
labeled objects (i.e. labeled graphs), while the atoms of a molecule
are of course not uniquely labeled. The mathematical concept to tackle
this problem is to consider orbits of labeled molecular graphs under
the operation of the symmetric group. However, when enumerating molecular
structures, one has thus to solve the so--called isomorphism problem, where molecules that seem different are actually equivalent (i.e., their atoms can be mapped on top of each other by keeping the connectivity between atoms intact).
The good news is that isomorphism can be solved in polynomial time for
molecular graphs due to the fact that molecules belong to a restricted
class of graphs known as \emph{bounded valence graphs}; however, that
same restriction prevents applying to a molecular graph the polynomial
delay enumeration results obtained for general graphs (Chapter 8 in
\cite{faulon2010}). While the complexity of molecular graph
enumeration remains an open problem it has nonetheless been shown that
molecules can be sampled efficiently \cite{goldberg1999}. Sampling
procedures based on the Metropolis or the Genetic Algorithms have been
developed and used to elucidate natural compounds from NMR data.

The underlying computational complexity and potential intractability
of molecular enumeration may not be a bottleneck after all, as
computational chemists have developed and successfully used
enumeration tools to generate large chemical libraries such as GDB-13,
comprising almost a trillion chemical
structures\footnote{http://www.dcb-server.unibe.ch/groups/reymond/gdb/home.html}. Yet,
the current enumeration software products do not generally produce
stereoisomers or tautomers; these require specific enumeration
procedures, which are still the subject of investigations by the
cheminformatics community. In addition, also GDB-13 only enumerates molecules up to 13 heavy atoms - given the possibly exponential increase in search space with the number of atoms, it seems unlikely that the practically relevant chemical space of about 30 heavy atoms will be exhaustively enumerated in the near future.

GDB-13 along with other databases such as PubChem (NCBI/NIH) or ZINC
(UCSF) are valuable resources to search for new drugs. For instance
the well known ligand-based design approach uses these databases along
with structure similarity tools to discover new compounds binding to
biological targets (protein or nucleic acid). Compounds retrieved
through ligand-based design, however, must already be present in
databases, which is a limitation of the technique. To overcome that
limitation, algorithms have recently been developed to enumerate all
compounds in the chemical space matching circular \cite{faulon2003} or
path fingerprints \cite{fujiwara2008}, these fingerprints being
popular tools to measure similarity. Enumerating molecules from
fingerprints, or more generally enumerating molecules from molecular
descriptors, is still very much under development. At stake is the
design of new chemicals or drugs that have not yet been synthesized.

Apart from analytical chemistry and compound design, enumeration plays a role is in the \emph{generation
  of chemical reaction networks}. The problem here consists of
enumerating all the possible compounds that can be produced by
applying reaction rules to a set of initial molecules. By reversing
the reaction rules one can also find sets of starting compounds
necessary for producing a given target. This latter process is known
in chemistry as \emph{retrosynthesis}, ever since the works of
E.J. Corey, who received the Nobel Prize in Chemistry in
1990. Designing new drugs or chemicals, understanding the kinetics of
combustion and petroleum refining, studying the dynamics of metabolic
networks, applying metabolic engineering and synthetic biology to
produce heterologous compounds in microorganisms, all are applications
that require the enumeration of reaction networks. As reviewed in
Chapter 11 in \cite{faulon2010} several network enumeration techniques
have been developed. However, these techniques generally suffer from a
combinatorial explosion of intermediate compounds being produced. One
way to limit the number of compounds generated is to simulate the
dynamics of the network while it is being constructed and remove
compounds of low concentration. Following that idea, methods have been
developed based on the Gillespie Stochastic Simulation Algorithm (SSA)
to compute on-the-fly species concentrations. Chemical reaction
network enumeration and sampling is an active field of research,
particularly in the context of metabolism, either to study
biodegradation, or to propose metabolic engineering strategies to
biosynthesize compounds of commercial interest. The difficulty with
metabolic network design is that in addition to network generation
based on reactions, one also needs to verify that there are enzymatic
events possible for enabling reactions catalyses. That additional task
requires encompassing both chemical and sequence information and the
development of tools that are at the interface between
cheminformatics and bioinformatics.

@@

The basis of predictive modeling in cheminformatics is that
\emph{chemical structure implies biological activity}. Thus, the goal
of any modeling approach is to capture and characterize the
correlations between structural features and the observed biological
activities. The goal is to describe cause-effect relationships with
the lowest possible risk, and also to describe the likelihood of error
(confidence) when using such models for decision making. In many
cases, the activity of a small molecule is due to its interaction with
a receptor. Traditional 'QSAR' approaches ignore receptor interactions,
focusing only on small molecule features, and therefore loose valuable
information. As a result, techniques such as proteochemometric
modeling have been designed to take into account both ligand and
receptor structures.

@@

>>>>

\usepackage{array}
\usepackage{pifont}
\usepackage{url}
\usepackage{graphicx}
\usepackage{multirow}

\newcommand{\none}{\ding{55}}
\newcommand{\least}{\ding{51}}
\newcommand{\little}{\ding{51}\ding{51}}
\newcommand{\lots}{\ding{51}\ding{51}\ding{51}}

@@

\maketitle
\begin{abstract}
One of the most prominent success stories in all the sciences over the last decade has
  been the advance of bioinformatics: The interdisciplinary collaboration between
 computer scientists and molecular biologists led as one of its major accomplishments to the sequencing of the human
 genome, where researchers connected biological concepts to the theory of string algorithms. However, despite this
 great
  success, few computer scientists are familiar with a related (and older!)
  discipline: cheminformatics, the use of computers to represent
 the structures of small molecules and analyze their properties.
  Cheminformatics has wide applicability, from the discovery of novel drug candidates to the optimization of
  certain physicochemical properties of small molecules. Until
 recently, the data and techniques employed in the cheminformatics field have been closely guarded
  secrets of companies whose financial success depended on being the first to produce
  the new ``miracle molecule.'' Only within the last decade---and as an effect of a change in mindsets,
  government-mandated policy changes and last but not least because of  chemists volunteering their time for an Open
  Science ``movement''---have researchers gained access to freely available software packages and databases of tens of
  millions of chemicals. Academic chemists now confront a variety of unsolved
  algorithmic problems that could not have been tackled a decade ago, but whose
  solutions are critical to research ranging from determining the behavior of small
  molecules in biological pathways, to finding a cure for malaria and other ``neglected'' diseases.
\end{abstract}

\category{J.2}{Computer Applications}{Physical Sciences and Engineering}[Chemistry]
\terms{Algorithms, Design, Human Factors, Theory}
\keywords{cheminformatics, chemoinformatics, graphs, chemistry, databases, knowledge management, open source, open
data}

@@

Novel technologies in the life sciences churn out information at an ever
increasing rate---public data stores such as the one at the European Bioinformatics
Institute (EBI) contain on the order of 10 petabytes of biological information.
However, while biological information has been publicly available for decades,
the same has not been true for chemical information, until recently. It was
not until 2004 that a large public small molecule structure repository (PubChem)
became freely available. This was followed by other databases, as discussed later
in this article. In the same vein, open source software and the publication of
algorithms have only recently become a main focus of the field
\cite{faulon2010}.

@@

\section{Bridging Cheminformatics and Computer Science}

@@

While cheminformatics has a broad applicability, ranging from
agrochemical research and drug discovery to the design of novel
materials (since everything around us consists of ``molecules''), we have chosen to use drug discovery as the context of
this
article due to its central relevance for human wellbeing. Given that the safety of drugs on the market is a key concern
both for the general audience as well as researchers in pharmaceutical companies, the remainder of this article is
structured around the
concept of ``risk minimization'' in drug discovery---the question of how to minimize the chances of a small molecule
failing in the various research and development stages as a drug candidate, and how to successfully employ
cheminformatics methods in these efforts.  Table~\ref{table:properties} (in the Appendix) summarizes properties a drug candidate must satisfy to be considered ``promising,'' and sketches the role of cheminformatics at each stage of drug development.
%
\subsection{Databases \& Ontologies}
\label{sec:databases}
%
Most cheminformatics applications rely on large databases of chemical structures,
their properties, and relationships to, \textit{e.g.}, biological targets.
Organizing and maintaining such databases, as well as searching and clustering similar structures together,
are essential  to enable many scientific applications. Yet, each one of
those areas poses its unique computer science challenges.

There is considerable progress in open chemical and bioactivity databases like
ChEMBL and NCBI-PubChem, which are both freely available and contain in the order of millions (ChEMBL) and tens of
millions (PubChem) of data points each. Still, the integration
of chemical databases with each other remains a challenge, not only due to the
normalization problems in the chemical and bioactivity landscape, but also due
to sheer data volume. One problem is the annotation with the ``protein targets'' a molecule binds to---here dozens of
protein and gene identifiers exist, generally without the availability of a one-to-one mapping from each ontology, which makes merging databases often a cumbersome exercise.

Given the requirement to ``make sense'' of as much chemical data at hand as possible, we have to be able to search in
complex data, for example multi-labelled molecular graphs, where graph labels can be different. Hence, we need to be
able to capture molecular graphs in a risk-bounded machine encoded way.

One main challenge in the cheminformatics field is that not one unique way exists to represent every molecule in a
machine-readable form. This results in multiple encodings of
``the same'' molecule and its different tautomeric or protonation states (see the Glossary in the Appendix for an explanation of technical chemistry terms). Checking whether two graphs represent the same chemical structure is
an expensive operation (the next section will discuss this in-depth);
thus, a representation of the structure that is unique and invariant
with respect to a particular graphical display of that chemical is needed for
identification. An early effort in this direction was the Simplified
Molecular Input Line Specification (SMILES), and specifically the
generation of a single, ``canonical'' SMILES for a particular structure. The idea was to devise rules for a unique
ordering of the atoms in a molecular structure, hence resulting in the same representation of the same molecule.
However,
over the course of time, several
different SMILES implementations became available, with no universal
standard for generating a canonical SMILES. Nonetheless, the SMILES format remains one of the major representations of
molecules used in practice today.

With each database using their own algorithm for molecular encoding,
automated exchange of data between different databases was hindered.
As more and more data became available online, the use of unique
structure-based identifiers became a pressing need, resulting in the IUPAC
International Chemical Identifier (InChI) to be developed. The
InChI is a non-proprietary, structured textual identifier for chemical
entities~\cite{inchi}. InChI identifiers are not intended to be read and
understood by humans, but are useful for computational matching of
chemical entities. For rapid database lookups, the InChIKey is a
hashed key for the InChI, which has an invariant length of 14
characters. (Unfortunately, after implementation and publication of the hashing scheme it became apparent that in
particular in case of stereochemically related molecules clashes of the InChI key take place in some, albeit rare
cases.) Figure~\ref{figure:smiles} illustrates the SMILES, InChI and InChIKey for
lipoic acid.

\begin{figure}
\centering
\includegraphics[height=1in]{lipoicacid.png}
\caption{The standard chemical graph illustration together with the SMILES, InChI and InChIKey codes for lipoic acid. Note that the SMILES is much more human-readable than the others.}
\label{figure:smiles}
\end{figure}

The InChI is now widely used for matching identical chemical structures, but it
is still limited. For example, it cannot differentiate between
certain types of stereoisomers: Figure~\ref{figure:cistrans} (in the Appendix) illustrates two stereoisomers for which
the generated InChI is the same.

@@

Graph substructure matching is a variant of \emph{graph isomorphism detection}, which is widely believed to be computationally
intractable~\cite{cordella2001}. To execute a graph isomorphism search
across a full chemical database of thousands or even millions of
structures is infeasible. For that reason, complex heuristics must be
used to drastically filter the candidate search space. Structure-based
\emph{fingerprints} are commonly used for this purpose. Fingerprints
encode characteristic features of a given chemical structure in a
Boolean array, or bitmap, and they can be implemented in a variety of
ways. Common fingerprints are created by an algorithm which generates
bitmap patterns for each atom in the structure; then each atom and its
nearest neighbors, including the bonds between them; then each group
of atoms connected by paths of increasing lengths up to the maximum
implemented for the algorithm (commonly around eight). In this way,
characteristic parts of molecule are captured and the bitmap patterns
are added together into the final fingerprint. If all bits in a query
fingerprint are also present in the target fingerprint of a stored
database structure, only then is the structure subjected to the
computationally expensive subgraph matching algorithm. Bit operations
are very fast and independent of the number of atoms in a structure
due to the fixed length of the fingerprint.

%Fingerprints also facilitate the rapid calculation of quantitative
%structure-based similarity measures. An example measure of similarity is the
%\emph{Tanimoto coefficient}, calculated as the ratio $T(a,b) = \frac{c}{(a+b-c)}$,
%where $c$ is the count of bits on in the same position in both the two
%fingerprints, $a$ is the count of bits on in the first structure, and $b$ is the
%count of bits on in the second structure. Historically, this coefficient originated from the comparison of ecosystems
%around lakes, where a measure was needed to describe how similar those ecosystems are with respect to the animals,
%plants etc. they contain. However, it can be applied to every other system as well where features can be described in a
%binary manner. The Tanimoto coefficient varies in the
%range 0.0 -- 1.0, with a score of 1.0 indicating that the two structures are either very
%similar or even identical (i.e. their fingerprints are the same).
%
Structure-based similarity is however of
limited use in grouping of chemicals based on non-structural features, such as
shared bioactivity profiles. For this purpose, a large-scale and flexible
classification structure is needed for chemicals and their linked biological
information. One such answer to this need is provided in the form of \emph{ontologies}.
An ontology is a formal specification of entities and their relationships in a
domain of interest. It is structured around an underlying logical representation
such as Description Logics~\cite{baaderdl2007}. The logic-based representation allows
automated reasoning to derive new knowledge (inferences) from the knowledge
encoded, making the representation both compact and powerful.
Ontologies are already in widespread use across the biomedical domain, and in the cheminformatics field ontologies
relating to the chemical structures themselves and the protein targets the molecules are active against are probably
most common. A
similarity measure based on the information encoded in ontologies is termed
\emph{semantic similarity}, and such measures have already been applied to enhance
chemical classification in several problem areas \cite{couto2010}. Many open
questions still remain about the best way to combine the information encoded in
chemical structures with chemical ontologies \cite{hastingsowled2010}.

@@

Enumerating molecules is a combinatorial problem that has fascinated
chemists, computer scientists and mathematicians alike, for more than a
century. It is one important risk-reduction element in still being
able to encode closely-related molecules and for keeping the space
complexity of databases under (algorithmic) control. Indeed, while
trying to solve the problem of counting the isomers of paraffin
structures or counting substituted aromatic compounds, fundamental
principles in graph theory and combinatorics were developed by Cayley,
Polya and others. Even terms that are widely-used nowadays, like ``graph''
and ``tree'', were originally coined in a chemistry context (Chapters 1 and
8 in \cite{faulon2010}). More than four decades ago, Joshua Lederberg,
Carl Djerassi, Edward Feigenbaum and others from Stanford University
started developing codes to explicitly enumerate molecules. Their
project, named DENDRAL, was funded by NASA with the goal of developing
an automated system to elucidate compounds from spectral data. While
studying this challenging problem, important concepts in computer
science were developed. DENDRAL is generally quoted in artificial
intelligence textbooks as a pioneer project that produced the first
expert system~\cite{DENDRAL}.

Enumerating molecules is not only an interesting academic exercise but has
practical utility as well. The foremost application of enumeration is \emph{structure
elucidation} of natural compounds such as metabolites. Metabolites are small molecules that can be found an organism
(such as a human) and which in many cases indicate the state the organism is in---such as indicating a disease, or a
fasted state, etc.---and with a sufficiently large database the idea is to be able to detect the chemical composition of
body fluids and to arrive at conclusions regarding the particular state of the organism at this point in time.
Ideally, the wishful bench chemist collects experimental data
from an unknown compound, the data is fed into a code, and the resulting
structure is returned. Although that streamlined process may not yield a unique
solution, there are commercial software products such as MOLGEN
(\url{http://molgen.de}) that can, for instance, list all structures matching a given
molecular formula, or mass spectrometry (MS) data. Another important application
of enumeration is \emph{molecular design}. Here the problem is to design compounds
(drugs, for instance) that optimize some physical, chemical, or biological
property or activity. Although less prolific and more recent than structure
elucidation, molecular design has introduced some novel solutions to molecular
enumeration. Finally, with the advent of \emph{combinatorial chemistry} (the rapid synthesis of different but related molecules), molecular
enumeration has taken a central role as it allows computational chemists to
construct virtual libraries and test hypotheses, and it provides guidance in the
design of optimal combinatorial experiments.

The major difficulty with enumeration (apart from the size of the search space, which is thought to be in the order of
$10^{60}$ molecules for 30 heavy atoms) is that the \emph{in silico}
representation of molecules, the so-called molecular graphs, are
labeled objects (i.e. labeled graphs), while the atoms of a molecule
are of course not uniquely labeled. The mathematical concept to tackle
this problem is to consider orbits of labeled molecular graphs under
the operation of the symmetric group. However, when enumerating molecular
structures, one has thus to solve the so--called isomorphism problem, where molecules that seem different are actually
equivalent (i.e., their atoms can be mapped on top of each other by keeping the connectivity between atoms intact).
The good news is that isomorphism can be solved in polynomial time for
molecular graphs due to the fact that molecules belong to a restricted
class of graphs known as \emph{bounded valence graphs}; however, that
same restriction prevents applying to a molecular graph the polynomial
delay enumeration results obtained for general graphs (Chapter 8 in~\cite{faulon2010}). While the complexity of molecular graph
enumeration remains an open problem it has nonetheless been shown that
molecules can be sampled efficiently~\cite{goldberg1999}. Sampling
procedures based on the Metropolis or the Genetic Algorithms have been
developed and used to elucidate natural compounds from NMR data.

The underlying computational complexity and potential intractability
of molecular enumeration may not be a bottleneck after all, as
computational chemists have developed and successfully used
enumeration tools to generate large chemical libraries such as GDB-13,
comprising almost a trillion chemical
structures~\cite{GDB}. Yet,
the current enumeration software products do not generally produce
stereoisomers or tautomers; these require specific enumeration
procedures, which are still the subject of investigations by the
cheminformatics community. In addition, GDB-13 only enumerates molecules up to 13 heavy atoms---given the possibly
exponential increase in search space with the number of atoms, it seems unlikely that the practically relevant chemical
space of about 30 heavy atoms will be exhaustively enumerated in the near future.

%GDB-13 along with other databases such as PubChem (NCBI/NIH) or ZINC
%(UCSF) are valuable resources to search for new drugs. For instance
%the well known ligand-based design approach uses these databases along
%with structure similarity tools to discover new compounds binding to
%biological targets (protein or nucleic acid). Compounds retrieved
%through ligand-based design, however, must already be present in
%databases, which is a limitation of the technique. To overcome that
%limitation, algorithms have recently been developed to enumerate all
%compounds in the chemical space matching circular \cite{faulon2003} or
%path fingerprints \cite{fujiwara2008}, these fingerprints being
%%popular tools to measure similarity. Enumerating molecules from
%fingerprints, or more generally enumerating molecules from molecular
%descriptors, is still very much under development. At stake is the
%design of new chemicals or drugs that have not yet been synthesized.
%
Apart from analytical chemistry and compound design, enumeration plays a role is in the \emph{generation
  of chemical reaction networks}. The problem here consists of
enumerating all the possible compounds that can be produced by
applying reaction rules to a set of initial molecules. By reversing
the reaction rules one can also find sets of starting compounds
necessary for producing a given target. This latter process is known
in chemistry as \emph{retrosynthesis}, ever since the works of
E.J. Corey, who received the Nobel Prize in Chemistry in
1990. Designing new drugs or chemicals, understanding the kinetics of
combustion and petroleum refining, studying the dynamics of metabolic
networks, applying metabolic engineering and synthetic biology to
produce heterologous compounds in microorganisms, all are applications
that require the enumeration of reaction networks. As reviewed in
Chapter 11 in \cite{faulon2010} several network enumeration techniques
have been developed. However, these techniques generally suffer from a
combinatorial explosion of intermediate compounds being produced. One
way to limit the number of compounds generated is to simulate the
dynamics of the network while it is being constructed and remove
compounds of low concentration. Following that idea, methods have been
developed based on the Gillespie Stochastic Simulation Algorithm (SSA)
to compute on-the-fly species concentrations. Chemical reaction
network enumeration and sampling is an active field of research,
particularly in the context of metabolism, either to study
biodegradation, or to propose metabolic engineering strategies to
biosynthesize compounds of commercial interest. The difficulty with
metabolic network design is that in addition to network generation
based on reactions, one also needs to verify that there are enzymatic
events possible for enabling reactions catalyses. That additional task
requires encompassing both chemical and sequence information and the
development of tools that are at the interface between
cheminformatics and bioinformatics.

@@

The basis of predictive modeling in cheminformatics is that
\emph{chemical structure implies biological activity}. Thus, the goal
of any modeling approach is to capture and characterize the
correlations between structural features and the observed biological
activities. The goal is to describe cause-effect relationships with
the lowest possible risk, and also to describe the likelihood of error
(confidence) when using such models for decision making.  Figure~\ref{figure:applicability-domain} (in the Appendix) illustrates the use of substructure matching to produce a space of chemical structures that are likely to share similar biological activity; this is called the generation of an \emph{applicability domain}.

In many cases, the activity of a small molecule is due to its interaction with
a receptor. Traditional ``QSAR'' approaches ignore receptor interactions,
focusing only on small molecule features, and therefore loose valuable
information. As a result, techniques such as proteochemometric
modeling have been designed to take into account both ligand and
receptor structures.

@@

\appendix

\section{Glossary of Terms}
\begin{itemize}
\item \textit{Bioassay} - A system measuring the biological activity of a chemical compound on a particular
    biological assay (with all its parameters).
\item \textit{Biodegradation} - is the chemical dissolution of materials by bacteria or other biological means.
\item \textit{Conformation and conformational isomerism} - In chemistry, conformational isomerism is a form of
    stereoisomerism in which the isomers can be interconverted exclusively by rotations about formally single
    bonds.
    Single possible states of compounds are called conformers.
\item \textit{Ligand (biochemistry)} - a substance that binds to a protein.
\item \textit{Ligand (coordination chemistry)} - an atom, ion, or functional group that donates one or more of its
    electrons through a coordinate covalent bond to one or more central atoms or ions.
\item \textit{Mass spectrometry} - Mass spectrometry (MS) is an analytical technique that measures the
    mass-to-charge ratio of charged particles. It is used for determining masses of particles, for determining the
    elemental composition of a sample or molecule, and for elucidating the chemical structures of molecules, such
    as
    peptides and other chemical compounds.
\item \textit{Metabolites} - Metabolites are the intermediates and products of metabolism. The term metabolite is
    usually restricted to small molecules (chemical compounds).
